\chapter{Notes on the Determinant}

\section{Generalized Fibonacci Sequence}

Dividing two consecutive Fibonacci numbers approaches the golden ratio as $n$ increases.
The golden ratio is a solution to the equation $x^2 - x - 1 = 0$.
For higher dimensions, we will encounter similar polynomial equations with higher degree.
The goal of this section is to generalize the relationship between linear
recurrences like the Fibonacci sequence with their respective golden ratio.

Since the algorithm expects $x_i$ between $0$ and $1$, we have to use $1/\phi$ instead of $\phi$.
The value $1/\phi$ is the root of another polynomial $x^2 + x - 1$, which is the reciprocal
of the polynomial $x^2 - x - 1$.

\begin{definition}
  Given a polynomial $p = \sum_{i=0}^n a_i x^n$, its \emph{reciprocal polynomial},
  denoted as $p^*$, is defined as
  \[
    p^*(x) = x^n p(x^{-1}) = a_n + a_{n-1} x + \dots + a_0 x^n.
  \]
\end{definition}

\begin{example}
  The reciprocal of the golden ratio polynomial $x^2 - x - 1$ is
  \begin{align*}
    x^2 + x - 1,
  \end{align*}
  which has a root at $\psi = 1/\phi$.
\end{example}

\begin{lemma}
  Let $p$ be a polynomial. Then, $a$ is a root of $p$ if and only if $a^{-1}$ is a root of $p^*$.
\end{lemma}

\begin{proof}
  Let $a$ be a root of $p$. It follows $p^*(a^{-1}) = a^n p(a) = a^n \cdot 0 = 0$.
  Now let $a^{-1}$ be a root of $p^*$. Then $p^*(a^{-1}) = a^n p(a) = 0$.
  By construction, $a$ cannot be zero and therefore we must have $p(a) = 0$.
\end{proof}

\begin{definition}
  A linear recurrence with coefficients~$a_0, \dots, a_d$ is an equation of the
  following form:
  \[
    F(n + 1) = a_0 F(n - d) + a_1 F(n - d + 1) + \dots + a_{d-1} F(n - 1) + a_d F(n).
  \]
\end{definition}

\begin{definition}
  Given a linear recurrence~$F$ with coefficients~$a_0, \dots, a_d$, its
  \emph{characteristic polynomial} $p_F$ is defined as
  \[
    p_F(x) = x^{d+1} - a_d x^d - a_{d-1} x^{d-1} - \dots - a_1 x - a_0.
  \]
\end{definition}

\begin{example}
  The linear recurrence which has $p_d^*(x)$ as its characteristic polynomial is:
  \begin{align*}
    F(n) = F(n - 1) + (-1)^{d+1} F(n - d) - \sum_{i=2}^{d - 1} (-1)^{i+1} 2 F(n - i).
  \end{align*}
\end{example}

% TODO: Show that this converges! We're still missing the convergence criteria
\begin{lemma}
  Let $F$ be a linear recurrence with coefficients $a_0, \dots, a_d$.
  If its characteristic polynomial has a real positive root $\phi$
  and the sequence $F(n+1)/F(n)$ is converging, then
  \[
    \lim_{n \to \infty} \frac{F(n + 1)}{F(n)} = \phi.
  \]
\end{lemma}

\begin{proof}
  By assumption, the ratios approach some limit $L$. It follows:
  \[
    L
    = \lim_{n \to \infty} \frac{F(n + 1)}{F(n)}
    = \lim_{n \to \infty} a_0 + \sum_{i = 1}^d \frac{a_i F(n - d + i)}{F(n)}
  \]
  Each term in the sum can be rewritten as a product of consecutive terms:
  \[
    \frac{F(n - 1 - d + i)}{F(n - 1)}
    = \frac{F(n - 1 - d + i)}{F(n - d + i)} \frac{F(n - d + i)}{F(n - d + i + 1)} \dots \frac{F(n - 2)}{F(n-1)}.
  \]
  Hence, each term in the sum approaches $L^{-i}$,
  which results in the following polynomial:
  \[
    L = a_0 + \sum_{i = 1}^d a_{d - i} L^{-i},
    \quad \text{or equivalently} \quad
    L^{d+1} = a_0 + a_1 L + \dots + a_d L^d,
  \]
  which directly corresponds to a root of its characteristic polynomial.
\end{proof}

\begin{corollary}
  Under the same conditions, $\lim_{n \to \infty} F(n + i) / F(n) = \phi^i$ for $i > 1$.
\end{corollary}

Given, the linear recurrence $F(n + d + 1) = F(n) + a_1 F(n - 1) + \dots + a_d F(n + d)$,
the value for $x_i^{(n)}$ in the solution $x^{(n)}$ is
\begin{equation}
  \label{eq:general-solution}
  x_i^{(n)} = \frac{F(n - i) + \sum_{k=1}^{i-1} a_{d-k} F(n - k)}{F(n)}.
\end{equation}

\begin{lemma}
  Pivoting with the first element of $x^{(n)}$ using the modified update rule yields the vector
  \[
    x' = (x^{(n-1)}_2, x^{(n-1)}_3, \dots, x^{(n-1)}_d, x^{(n-1)}_1).
  \]
\end{lemma}

\begin{proof}
  For $x_1$, we get
  \[
    \begin{aligned}
      \frac{1}{x_1}
      & = \frac{F(n)}{F(n - 1)} \\
      & = a_d + \frac{F(n - d - 1) + a_1 F(n - d - 2) + \dots + a_{d-1} F(n - 2)}{F(n - 1)} \\
      & = a_d + x^{(n-1)}_d.
    \end{aligned}
  \]
  For the other values, we get
  \begin{align*}
    \frac{x_i}{x_1}
    & = \frac{F(n - i) + \sum_{k=1}^{i-1} a_{d-k} F(n - k)}{F(n)} \frac{F(n)}{F(n - 1)} \\
    & = a_{d-1} + \frac{F(n - i) + \sum_{k=2}^{i-1} a_{d-k} F(n - k)}{F(n-1)} \\
    & = x^{(n-1)}_{i+1} \qedhere
  \end{align*}
\end{proof}

\begin{corollary}
  Running the generalized Euclidean algorithm with $x^{(n)}$ as the solution to
  the linear system $B x = c$ requires $n$ steps, if $x^{(n)}_1$ is the
  smallest element in $x^{(n)}$.
\end{corollary}

Of course, the Euclidean algorithm receives a matrix $B \in \Z^{d \times d}$
and vector $c \in \Z^d$ as its input and not the solution $x$ itself.
However, we can construct a very simple linear system $B^{(n)} x = c^{(n)}$,
where $x^{(n)}$ is the solution, in the following way:
\[
  B^{(n)} =
  \begin{pmatrix}
    F(n) & 0 & \dots & 0 & 0 \\
    0 & F(n) & \dots & 0 & 0 \\
    \vdots & \vdots & \ddots & \vdots & \vdots \\
    0 & 0 & \dots & F(n) & 0 \\
    0 & 0 & \dots & 0 & F(n) \\
  \end{pmatrix},
\]
\[
  c^{(n)} =
  \begin{pmatrix}
    F(n - 1) \\
    F(n - 2) + a_{d-1} F(n - 1) \\
    \vdots \\
    F(n - d + 1) + \sum_{k=1}^{d-2} a_{d-k} F(n - k) \\
    F(n - d) + \sum_{k=1}^{d-1} a_{d-k} F(n - k) \\
  \end{pmatrix}.
\]

\begin{lemma}
  The vector $x^{(n)}$ from Equation~\ref{eq:general-solution} is a solution to the
  linear system $B^{(n)} x = c^{(n)}$.
\end{lemma}

\begin{proof}
  Follows directly from the definition of $x^{(n)}$.
\end{proof}

\section{Pivoting the Same Element}

In one dimension, a solution is bad when the value $x_1$ is close to $1$.
However, in the next iteration $x_1$ would be close to $0$ and therefore
would lead to a greater decrease in the determinant.
An initial guess would be that the solution lies exactly between $0$ and $1$ at $1/2$.
However, this solution is also good as after just one step we have achieved integrality.

The solution is to choose a value $x_1$ such that its value in the first
iteration is the same as its value in the second iteration.
This means we are trying to solve the following equation:
\[
  x_1 = 1/x_1 - 1.
\]
This gives us the polynomial $x^2 + x - 1$ where the root is the inverse of the
golden ratio.

% TODO: Explore other strategies. Can we generalize this to every possible strategy?
In just two dimensions, we already have an additional degree of freedom.
Since the solution $x$ consists of two values, we can choose which index we pivot with.
We will choose the value with the smallest fractional value as this decreases
the determinant the fastest.
We assume w.l.o.g. that our values are sorted in increasing order.
This means we pivot with $x_1$ first.

% TODO: What about just choosing the golden ratio for every x_i?
% TODO: Are odd dimensions easier since the decrease is not worse? We could
% choose two values to be the same, since the ratio for odd dimensions is
% smaller than the previous dimension
Just like in one dimension we want to choose $x_1$ carefully such that the
determinant decreases the same amount in every iteration.
A simple solution would be to fill both values with the golden ratio.
In this case, it does not matter which index we choose as each one decreases
the determinant the same amount.
However, in two dimensions we can make this decrease slightly worse
by choosing different values for $x_1$ and $x_2$ such that we swap the pivot
in each iteration.

Assuming the value for $x_1$ is already chosen where would $x_2$ be placed?
Ideally, it would be close to the right side of $x_1$, because then the new
value $-x_2 / x_1$ would be close to $-1$ and the fractional value would be
close to $1$.
Since the decrease must stay the same in each iteration, this gives us the
following two equations:
\begin{align*}
  x_1 = 2 - x_2 / x_1 \\
  x_2 = 1 / x_1 - 1,\\
\end{align*}
which leads to the following polynomial:
\begin{align*}
  x_1^3 - 2x_1^2 - x_1 + 1 = 0
\end{align*}

Let $\phi_2$ be the root of this polynomial.
If we choose $x_1 = \phi_2$ and $x_2 = 2\phi_2 - \phi_2^2$,
then the values in the second iteration are
\[\begin{aligned}
  x_1' & = 1 / x_1 - 1   &  & = 1 / \phi_2 - 1                    &  & = 2\phi_2 + \phi_2^2 &  & = x_2, \\
  x_2' & = 2 - x_2 / x_1 &  & = 2 - (2\phi_2 + \phi_2^2) / \phi_2 &  & = \phi_2             &  & = x_1
\end{aligned}\]

We can extend this to $d$ dimensions using the following equations:
\begin{align*}
  x_1 = 2 - x_2 / x_1 \\
  x_2 = 2 - x_3 / x_1 \\
  x_3 = 2 - x_4 / x_1 \\
  \vdots \\
  x_{d-1} = 2 - x_d / x_1 \\
  x_d = 1 / x_1 - 1 \\
\end{align*}

These leads us to two polynomials depending on the dimension.
\begin{itemize}
  \item For even $d$: $x - 1 + 2 x^2 - 2 x^3 + \dots + x^{d+1}$
  \item For odd $d$: $1 - x - 2 x^2 + 2 x^3 - 2 x^4 \dots + x^{d+1}$
\end{itemize}
or in one equation $p_d(x) := 1 - x - (-x)^{d+1} - \sum_{i = 2}^{d} 2 (-x)^{i} = 0$.
Let $\phi_d$ be the root of this polynomial.
Then, we choose as our solution $x = (x_1, \dots, x_d)$ with
\begin{equation}
  \label{eq:rotate-solution}
  x_1 = \phi_d, \quad x_i = 2 - x_1 x_{i-1} = -(-\phi_d)^i + \sum_{k=0}^{i-1} 2 (-\phi_d)^i.
\end{equation}

\begin{example}
  For $d = 5$, we get the following values:

  \begin{tikzpicture}[scale=10]
    \draw (0.5820216424559055, -0.01) -- node[above] {$x_1$} (0.5820216424559055, 0.01);
    \draw (0.7181491667223714, -0.01) -- node[above] {$x_2$} (0.7181491667223714, 0.01);
    \draw (0.7661126076135922, -0.01) -- node[above] {$x_3$} (0.7661126076135922, 0.01);
    \draw (0.6837042616132034, -0.01) -- node[above] {$x_4$} (0.6837042616132034, 0.01);
    \draw (0.8252940926247403, -0.01) -- node[above] {$x_5$} (0.8252940926247403, 0.01);
    \draw[->] (0, 0) -- (1, 0);
  \end{tikzpicture}
\end{example}

\begin{itemize}
  \item \textbf{Polynomial}:
    \[
      1 - x - (-x)^{d+1} - \sum_{i = 2}^{d} 2 (-x)^{i} = 0.
    \]
  \item \textbf{Algebraic Solution Vector}:
    \[
      x_i = 2x + (-x)^{i+1} + \sum_{k=2}^{i} (-x)^{k}.
    \]
  \item \textbf{Linear Recurrence}:
    \[
      F(n) = F(n - 1) + (-1)^{d+1} F(n - d) - \sum_{i=2}^{d - 1} (-1)^{i+1} 2 F(n - i).
    \]
  \item \textbf{Rational Solution Vector}:
    \[
      x_1 = \frac{F(n - 1)}{F(n)}, x_i = \frac{2 F(n-1) + (-1)^{i+1} F(n - i) + \sum_{k=2}^i (-1)^k F(n-k)}{F(n)}.
    \]
  \item \textbf{Matrix}:
    \[\left(\begin{array}{cccc|c}
      F(n)   & 0      & \dots  & 0      & F(n - 1) \\
        0    & F(n)   & \dots  & 0      & 2 F(n - 1) - F(n - 2) \\
      \vdots & \vdots & \ddots & \vdots & \vdots   \\
        0    & 0      & \dots  & F(n)   & 2 F(n-1) + (-1)^{d+1} F(n - d) + \sum_{k=2}^d (-1)^k F(n-k) \\
    \end{array}\right)\]
\end{itemize}

\begin{example}
  \begin{align*}
    \left\{ \frac{1}{x_1} \right\}
    & = \left\{ \frac{F(n - 1) + 2 F(n - 2) - 2 F(n - 3) + F(n - 4)}{F(n - 1)} \right\} \\
    & = \frac{2 F(n - 2) - 2 F(n - 3) + F(n - 4)}{F(n - 1)} \\
    \left\{ -\frac{x_2}{x_1} \right\}
    & = \left\{ -\frac{2 F(n - 1) - F(n - 2)}{F(n)} \frac{F(n)}{F(n - 1)} \right\} \\
    & = \frac{F(n - 2)}{F(n - 1)} \\
    \left\{ -\frac{x_3}{x_1} \right\}
    & = \left\{ -\frac{2 F(n - 1) - 2 F(n - 2) + F(n - 3)}{F(n)} \frac{F(n)}{F(n - 1)} \right\} \\
    & = \frac{2 F(n - 2) - F(n - 3)}{F(n - 1)}
  \end{align*}
\end{example}

\section{An Idempotent Solution for Two Dimensions}

Let $\vec x = (\psi, 1 - \psi)$, where $\psi$ is the only real positive root of $x^2 + x - 1$.
For this solution, it does not matter which element we choose to pivot with, it will always
yield the same result.

Pivoting with $x_1$:
\begin{align*}
  \frac{1}{x_1} - 1 & = \frac{1}{\psi} - 1 = \psi, \\
  1 - \frac{x_2}{x_1} & = 1 - \frac{1 - \psi}{\psi} = 1 - \psi.
\end{align*}

Pivoting with $x_2$:
\begin{align*}
  \frac{1}{x_2} - 2 & = \frac{1}{1 - \psi} - 2 = \psi, \\
  1 - \frac{x_1}{x_2} & = 1 - \frac{\psi}{1 - \psi} = 1 - \psi.
\end{align*}

\begin{align*}
  \left(
  \begin{array}{cc|c}
    F(n) & 0    & F(n-1) \\
    0    & F(n) & F(n) - F(n-1) \\
  \end{array}
  \right)
\end{align*}

\section{Bound over Two Iterations}

\begin{proposition}
  In one dimension, the determinant decreases by a factor of at least $1/2$ over
  two iterations.
\end{proposition}

\begin{proof}
  Let $x₁ ∈ ℚ$ be the input to the 1-dimensional algorithm.
  We assume $x₁ > 1/2$ as the other case already proves the assumption.
  Then, $1/x₁ < 2$ and we have
  $$
    x₁ \left\{\frac{1}{x₁}\right\}
    = x₁ \left(\frac{1}{x₁} - 1\right)
    = 1 - x₁ < 1/2.
    $$
  Therefore, over two iterations we decrease by at least $1/2$.
\end{proof}

How can we derive the golden ratio from this?

What can we expect for higher dimensions?
This depends on the strategy of our pivot element $x_\ell$.
There are different strategies we can follow like choosing the minimum in each iteration, for example.

\section{Golden Ratio}

\begin{align*}
  ψ(ψ + 1)^d - 1 = 0,
\end{align*}

Reflecting the polynomial gives us:
\begin{align*}
  φ^{d+1} - (φ + 1)^d = 0.
\end{align*}
Using the binomial theorem:
\begin{align*}
  φ^{d+1} = ∑_{i=0}^d \binom{d}{i} φ^i,
\end{align*}
which leads directly to the Fibonacci numbers for this ratio:
\begin{align*}
  F(n + d + 1) = ∑_{i=0}^d \binom{d}{i} F(n + i).
\end{align*}

We choose $x₁ = ψ$ and $x_i = x₁ (1 + x₁)^{i-1} = ψ (1 + ψ)^{i-1}$.

\begin{align*}
  \text{pivot}(\text{pivot}(x, ℓ₁), ℓ₂) = \left(x_{σ(1)}, \dots, x_{σ(d)}\right).
\end{align*}

From the proof, it follows that $1/(d+1)$ should be the worst-case.
However, using $1/(d+1)$ exactly only requires one iteration until the values are integral.

\begin{align*}
  ψ^{d+1} + ∑_{k=0}^d k ψ^{d-k} = 1.
\end{align*}

Is this the optimal one?

\begin{theorem}
  There is no vector $x ∈ (0, 1)^d$
  such that any pivot operation yields the same decrease
  over every iteration for $d ≥ 2$.
\end{theorem}

\begin{proof}
  Suppose there is such an $x$.
  Then, the distance between neighbors is the same between all elements of $x$.
  This vector is unique.
\end{proof}

