\chapter{The Euclidean Algorithm}

\section{Analysis}

\begin{Pseudocode}
algorithm Euclidean(a, b)
  while $b ≠ 0$ do
    $(a, b) ← (b, a \bmod b)$
  end

  return $a$
end
\end{Pseudocode}

\begin{proposition}
  The ratio $b/a$ decreases by at least $1/2$ over two iterations.
\end{proposition}

\begin{proof}
  Suppose $b/a > 1/2$ in the first iteration.
  Over two iterations, we have
  \[
    \frac{b}{a} · \frac{a \bmod b}{b}
    ≤ \frac{b}{a} · \frac{a - b}{b}
    = 1 - \frac{b}{a}
    < 1 - \frac{1}{2}
    = \frac{1}{2}.
  \]
  Therefore, it decreases by at least $1/2$ over two iterations.
\end{proof}

\section{Fibonacci Numbers}

\begin{definition}
  The \emph{Fibonacci numbers} are defined as
  \begin{enumerate}
    \item $F(0) = F(1) = 1$.
    \item $F(n) = F(n - 1) + F(n - 2)$ for $n > 2$.
  \end{enumerate}
\end{definition}

\begin{proposition}[Lamé's Theorem \cite{Lame1844}]
  The Euclidean algorithm requires at most $5h$ steps,
  where $h = \log_{10}(b)$.
\end{proposition}

\begin{proof}
  \todo[inline]{Using Fibonacci numbers.}
\end{proof}

\section{Extension to the Real Numbers using Continued Fractions}

What if the ratio $a/b$ perfectly equals the golden ratio?
In this case, the Euclidean algorithm no longer runs on two integer inputs,
since the golden ratio is irrational.
Therefore, we extend the algorithm to the real numbers.
Our input now consists of a single number $x ∈ ℝ$, which in the original
algorithm represents the ratio between the two inputs.
This allows us to run the algorithm on the golden ratio.

Given $x ∈ ℝ$, the algorithm proceeds just as the original.
We assume $x$ actually represents the fraction $x/1$,
then we find a number $a₀ ∈ ℕ$ times until $x - 1 · a₀ < 1$.
Once this is the case, we swap the inputs, i.e. we calculate the inverse fraction $1/(x - a)$.
Now, we subtract $x - a$ from $1$ until $1 - a₁ (x - a₀) < x - a₀$
and then calculate the inverse fraction again.
We repeat this process until $x = 0$.
Subtracting $a_i$ from the numerator such that it is smaller than the
denominator and then calculating the inverse.

\begin{example}
  Let $x = 13/5$.
  The algorithm proceeds as follows:
  \[
    \begin{array}{rclcrcl}
      13/5 & = & 2 & · & 1   & + & 3/5 \\
         1 & = & 1 & · & 3/5 & + & 2/5 \\
       3/5 & = & 1 & · & 2/5 & + & 1/5 \\
       2/5 & = & 2 & · & 1/5 & + & 0.
    \end{array}
  \]
\end{example}

\begin{example}
  Let $x$ be the golden ratio.
  The algorithm proceeds as follows:
  \[
    \begin{array}{rclcrcl}
      φ & = & 1 & · & 1   & + & φ - 1 \\
         1 & = & 1 & · & (φ - 1) & + & (2 - φ) \\
       φ - 1 & = & 1 & · & (2 - φ) & + & (2φ - 3) \\
       2 - φ & = & 1 & · & (2φ - 3) & + & (5 - 3φ) \\
       & \vdots &
    \end{array}
  \]
\end{example}


The real algorithm computes the coefficients of the continued fractions representation of $x$.
Continued fractions are fractions of the form
\[
  x = a_0 + \cfrac{1}{a_1 + \cfrac{1}{a_2 + \cfrac{1}{a_3 + \cfrac{1}{a_4 + \ddots}}}}.
\]
A continued fraction is finite, when its sequence $a_n$ is finite.
In this case, the continued fraction represents a rational number.

\section{Periodicity and Quadratic Irrationals}

\begin{definition}
  A continued fraction $[a₀; a₁, …]$ is called \emph{eventually periodic}
  if there exists an index $k₀ ≥ 0$ and a period $ℓ ≥ 1$ such that $aₖ = a_{k+ℓ}$ for every $k ≥ k₀$.
  A continued fraction is called \emph{purely periodic} if $k₀ = 0$.
\end{definition}

\begin{definition}
  A number $x ∈ ℝ$ is said to be a \emph{quadratic irrational} if $x$ is the root of some
  polynomial $p(x)$ over the rational numbers with degree $2$.
\end{definition}

\begin{example}
  Let $x = \sqrt{7}$.
  The algorithm proceeds as follows:
  \[
    \begin{array}{rclcrcl}
      \sqrt{7}      & = & 2 · 1               + (\sqrt{7} - 2)    & ≈ & 2.6458 & = & 2 · 1 + 0.6458      \\
      1             & = & 1 · (\sqrt{7} - 2)  + (3 - \sqrt{7})    & ≈ & 1      & = & 1 · 0.6458 + 0.3542 \\
      \sqrt{7} - 2  & = & 1 · (3 - \sqrt{7})  + (2\sqrt{7} - 5)   & ≈ & 0.6458 & = & 1 · 0.3542 + 0.2915 \\
      3 - \sqrt{7}  & = & 1 · (2\sqrt{7} - 5) + (8 - 3\sqrt{7})   & ≈ & 0.3542 & = & 1 · 0.2915 + 0.0627 \\
      2\sqrt{7} - 5 & = & 4 · (8 - 3\sqrt{7}) + (14\sqrt{7} - 37) & ≈ & 0.2915 & = & 4 · 0.0627 + 0.0407 \\
      & \vdots & & \vdots & & \vdots &
    \end{array}
  \]
\end{example}

The interesting connection between the quadratic irrationals and continued
fraction is that a continued fraction of a number $x ∈ ℝ$ is eventually
periodic if and only if $x$ is a quadratic irrational.
We begin with the first direction:

\begin{lemma}
  Let $x ∈ ℝ$, then
  \[
    [a₀; a₁, …, a_n, x] = [a₀; a₁, …, a_n + 1/x]
  \]
\end{lemma}

\begin{proof}
  \label{lem:nesting}
  If $n = 0$, then
  \[
    [a₀; x] = a₀ + \frac{1}{[x]} = a₀ + \frac{1}{x} = [a₀ + 1/x].
  \]
  Suppose the lemma is true for some $n ≥ 0$, then
  \begin{align*}
    [a₀; a₁, …, aₙ, x]
    & = a₀ + \frac{1}{[a₁; a₂, …, aₙ, x]} \\
    & = a₀ + \frac{1}{[a₁; a₂, …, aₙ + 1/x]} \\
    & = [a₀; a₁, …, aₙ, x]. \qedhere
  \end{align*}
\end{proof}

We define the following sequences over the sequence $\{a_n\}$:
\begin{align*}
  pₙ & = p_{n-1} a_n + p_{n - 2}, & p_{-1} & = 1, & p_{-2} & = 0, \\
  qₙ & = q_{n-1} a_n + q_{n - 2}, & q_{-1} & = 0, & q_{-2} & = 1.
\end{align*}

\begin{lemma}
  \label{lem:wallis}
  Let $x ∈ ℝ$, then
  \[
    [a₀; a₁, …, a_{n-1}, x] = \frac{p_{n-1} x + p_{n-2}}{q_{n-1} x + q_{n-2}}.
  \]
\end{lemma}

\begin{proof}
  If $n = 0$, then
  \[
    [x] = x = \frac{1x + 0}{0x + 1} = \frac{p_{-1} x + p_{-2}}{q_{-1} x + q_{-2}}.
  \]
  Suppose, the lemma is true for $n ≥ 0$.
  By Lemma~\ref{lem:nesting}, we have
  \begin{align*}
    [a₀; a₁, …, aₙ, x]
    & = [a₀; a₁, …, aₙ + 1/x].
  \end{align*}
  From the induction hypothesis, it follows that
  \begin{align*}
    [a₀; a₁, …, aₙ + 1/x]
    & = \frac{p_{n - 1} (aₙ + 1/x) + p_{n-2}}{q_{n-1} (aₙ + 1/x) + q_{n-2}} \\
    & = \frac{x (p_{n-1} aₙ + p_{n-2}) + p_{n-1}}{x (q_{n-1} aₙ + q_{n-2}) + q_{n-1}} \\
    & = \frac{x pₙ + p_{n-1}}{x pₙ + p_{n-1}}. \qedhere
  \end{align*}
\end{proof}

\begin{theorem}
  If the continued fraction representation of a number $x ∈ ℝ$ is eventually periodic,
  then $x$ is a quadratic irrational.
\end{theorem}

\begin{proof}
  Let $x$ be a continued fraction $[a₀; a₁, …]$ with a period of length $ℓ ≥ 1$
  starting at an index $k ≥ 0$,
  i.e. $x_k = [a_k; a_{k+1}, …] = [a_{k+ℓ}; a_{k+ℓ+1}, …] = x_{k+ℓ}$.
  By Lemma~\ref{lem:wallis}, we have
  \[
    x
    = \frac{p_k x_k + p_{k-1}}{q_k x_k + q_{k-1}}
    = \frac{p_{k+ℓ} x_{k+ℓ} + p_{k+ℓ-1}}{q_{k+ℓ} x_{k+ℓ} + q_{k+ℓ-1}}
    = \frac{p_{k+ℓ} x_k + p_{k+ℓ-1}}{q_{k+ℓ} x_k + q_{k+ℓ-1}}
  \]
  Multiplying by the denominators on both sides results in the quadratic equation
  \[
    (q_{k+ℓ} x_k + q_{k+ℓ-1})(p_k x_k + p_{k-1}) - (q_k x_k + q_{k-1}) (p_{k+ℓ} x_k + p_{k+ℓ-1}) = 0.
  \]
  Hence, $x_k$ is a quadratic irrational and $x$ must be one too since $x ∈ ℚ(x_k)$.
\end{proof}

The converse was originally shown by Lagrange.
Here I present a proof based on \cite{Northshield11},
which is much shorter and easier to understand.
Let $x$ be a quadratic irrational with polynomial $p$.
Furthermore, let $x = [a₀; a₁, \dots]$ be its continued fraction and $x_k = [a_k; a_{k+1}, \dots]$.
For every $x_i$ we construct a polynomial $p_i$ with $x_i$ as its root.
We begin by expanding $x$ once as $a₀ + x₁^{-1}$.
By definition, $a₀ + x₁^{-1}$ is the root of $p$.
However, this gives us a new polynomial in $x₁$ (if we multiply by $x₁^2$):
\begin{align*}
  p₁(x₁) = x₁^2 p(a_0 + x_1^{-1}) = A x₁^2 (a₀ + x₁^{-1})^2 + B x₁^2 (a₀ + x₁^{-1}) + C x₁^2 = 0. \\
\end{align*}
Reordering the polynomial into the standard quadratic form results in
\begin{align*}
  p₁(x₁) = \underbrace{(A a₀^2 + B a₀ + C)}_{A₁} + \underbrace{(2A a₀ + B)}_{B₁} + \underbrace{A}_{C₁}.
\end{align*}

We can continue this process for every $i$.
In particular, the coefficients of each polynomial $p_{i+1}$ is calculated
using the following recurrence relation:
\begin{align*}
  A_{i+1} = A_i a_i^2 + B_i a_i + C_i, \qquad
  B_{i+1} = 2 A_i a_i + B_i, \qquad
  C_{i+1} = A_i.
\end{align*}

The goal is to show that these coefficients are bounded in some way.
It follows that there must be some triple which repeats and therefore the
representation must be periodic.
The first step towards this proof is bounding the discriminant of each
polynomial $pᵢ$.

\begin{lemma}
  \label{lem:same-disc}
  The polynomials $p₁, p₂, …$ have the same discriminant.
\end{lemma}

\begin{proof}
  Let $Δ$ be the discriminant of $p₁$.
  By induction, suppose $p_i$ has discriminant $Δ$.
  Then the polynomial $p_{i+1}$ has discriminant
  \begin{align*}
    \phantom{= {}} B_{i+1}^2 - 4 A_{i+1} C_{i+1}
    & = (2 A_i a_i + B_i)^2 - 4 (A_i a_i^2 + B_i a_i + C_i) A_i \\
    & = 4 A_i^2 a_i^2 + 4 A_i B_i a_i + B_i^2 - 4 A_i^2 a_i^2 - 4 B_i A_i a_i - 4 C_i A_i \\
    & = B_i^2 - 4 A_i C_i \\
    & = Δ. \qedhere
  \end{align*}
\end{proof}

In order for the continued fraction to be periodic, the coefficients $A_i, B_i, C_i$ must be bounded.
The only way this could not be the case is when $A_i C_i > 0$,
since then there would be infinitely many values for such that $B_i^2 - 4 A_i C_i$.
However, this is not the case.

\begin{lemma}
  \label{lem:infinite-neg}
  There exists infinitely many polynomials $p_i$ with $A_i C_i < 0$.
\end{lemma}

\begin{proof}
  Suppose $A_i C_i > 0$ for every $i$ after some point.
  We assume WLOG that $A_i > 0$ since the negation of $p_i$ has the same roots.
  Except in the first iteration, every $x_i$ is positive and therefore $B_i$ must be negative.
  But $B_i = 2 A_{i-1} a_{i-1} + B_{i-1}$ would be strictly increasing since
  $A_i$ and $a_i$ are both assumed to be positive.
  This is a contradiction and therefore there must be infinitely many $p_i$ with $A_i C_i < 0$.
\end{proof}

\begin{theorem}[Lagrange's Theorem]
  If $x$ is a quadratic irrational, then its continued fraction is eventually periodic.
\end{theorem}

\begin{proof}
  By Lemma~\ref{lem:same-disc}, every polynomial~$p_i$ has the same
  discriminant~$Δ = B_i^2 - 4 A_i C_i > 0$ independent of the index~$i$
  and by Lemma~\ref{lem:infinite-neg} there are infinitely many polynomials
  such that $B_i^2$ and $-4 A_i C_i$ are both positive.
  But there can only be finitely many positive integers which sum up to another positive integer.
  Hence, there must be some triple $(A_i, B_i, C_i)$ which repeats, i.e. there
  exists an integer $i ≠ j$ with $(A_i, B_i, C_i) = (A_j, B_j, C_j)$.
  These polynomials must have the same root $x_i = x_j$ and by construction of
  the continued fraction, we have $a_{i+ℓ} = a_{j+ℓ}$ for every $ℓ ≥ 0$.
\end{proof}
