\chapter{Higher-Order Fibonacci Numbers and their Golden Ratios}
\label{ch:fibonacci}

% TODO: Add reference to Fibonacci numbers in the continued fraction chapter
One of the simplest periodic continued fraction is $[1; \overline{1}]$.
This fraction evaluates to the golden ratio and its convergents are ratios of
consecutive Fibonacci numbers.
Naturally, we can consider the simplest vectors $x ∈ ℝ^d$ for the generalized
Euclidean algorithm, where the integer part in each iteration is $1$.
Such vectors can be seen as a generalization of the golden
ratio to higher dimensions and their approximation can be seen as a
generalization of Fibonacci numbers.
However, there does not exist a single definitive golden ratio,
Since the generalized algorithm allows an additional freedom in the choice of the pivot.
So there are actually multiple different possible definitions.
Some of these are well known, like the supergolden ratio or the plastic ratio,
which are both roots of cubic polynomials.

Since the definition of the golden ratios depends entirely on the choice of our
pivot element $x_ℓ$, we consider two strategies for choosing $x_ℓ$ in this
chapter: First choosing the smallest fractional value $\{x_ℓ\}$ and second
choosing the largest fractional value $\{x_ℓ\}$.
Both of these strategies have a respective golden ratio for which the algorithm
becomes periodic and they each have a corresponding linear recurrence, which
can be seen as a generalization of the Fibonacci numbers.
In the end, we will derive a general definition of higher-order Fibonacci
sequences and show that the generalized Euclidean algorithm is periodic for
their golden ratios.

% ==============================================================================
\section{The Classical Case: Fibonacci Numbers and the Golden Ratio}
% ==============================================================================

\begin{definition}
  The \emph{Fibonacci numbers} are defined as
  \begin{enumerate}
    \item $F(0) = F(1) = 1$.
    \item $F(n) = F(n - 1) + F(n - 2)$ for $n > 2$.
  \end{enumerate}
\end{definition}

\begin{proposition}[Lamé's Theorem \cite{Lame44}]
  The Euclidean algorithm requires at most $5h$ steps,
  where $h = \log_{10}(b)$.
\end{proposition}

\begin{proof}
  \todo[inline]{Using Fibonacci numbers.}
\end{proof}

What if the ratio $a/b$ perfectly equals the golden ratio?
In this case, the Euclidean algorithm no longer runs on two integer inputs,
since the golden ratio is irrational.
Therefore, we extend the algorithm to the real numbers.
Our input now consists of a single number $x ∈ ℝ$, which in the original
algorithm represents the ratio between the two inputs.
This allows us to run the algorithm on the golden ratio.

Given $x ∈ ℝ$, the algorithm proceeds just as the original.
We assume $x$ actually represents the fraction $x/1$,
then we find a number $a₀ ∈ ℕ$ times until $x - 1 · a₀ < 1$.
Once this is the case, we swap the inputs, i.e. we calculate the inverse fraction $1/(x - a)$.
Now, we subtract $x - a$ from $1$ until $1 - a₁ (x - a₀) < x - a₀$
and then calculate the inverse fraction again.
We repeat this process until $x = 0$.
Subtracting $a_i$ from the numerator such that it is smaller than the
denominator and then calculating the inverse.

\begin{example}
  Let $x = 13/5$.
  The algorithm proceeds as follows:
  \[
    \begin{array}{rclcrcl}
      13/5 & = & 2 & · & 1   & + & 3/5 \\
         1 & = & 1 & · & 3/5 & + & 2/5 \\
       3/5 & = & 1 & · & 2/5 & + & 1/5 \\
       2/5 & = & 2 & · & 1/5 & + & 0.
    \end{array}
  \]
\end{example}

\begin{example}
  Let $φ$ be the golden ratio,
  i.e. $φ^2 - φ - 1 = 0$.
  The algorithm proceeds as follows:
  \[
    \begin{array}{rclcrcl}
      φ & = & 1 & · & 1   & + & φ - 1 \\
         1 & = & 1 & · & (φ - 1) & + & (2 - φ) \\
       φ - 1 & = & 1 & · & (2 - φ) & + & (2φ - 3) \\
       2 - φ & = & 1 & · & (2φ - 3) & + & (5 - 3φ) \\
       & \vdots &
    \end{array}
  \]
\end{example}

% ==============================================================================
\section{Higher-Order Fibonacci Numbers}
% ==============================================================================

We have already seen that the worst-case input for the one-dimensional
Euclidean algorithm are the Fibonacci numbers.
Plugging in two Fibonacci numbers always requires at least n steps for the
algorithm.
This leads to the question whether such numbers also exist in higher dimensions
and what their golden ratios look like.

% TODO: There should be a discussion on what it even means to have a Fibonacci number in higher dimensions

First, we have to understand what Fibonacci numbers in higher dimensions would
look like.
Fibonacci numbers have two properties which make them special for the Euclidean algorithm.
The first is that applying one iteration of the Euclidean algorithm on two
consecutive Fibonacci numbers~$(F(n + 1), F(n))$ yields the previous Fibonacci numbers~$(F(n), F(n - 1))$.
So the Euclidean algorithm requires exactly $n$ steps on this input.

The second property is that they are minimal with respect to this.
There are other sequences, which fulfill the first property, but the Fibonacci
numbers are special, in that they fit exactly once inside the next number.
This means that in the Euclidean algorithm there is a multiplicative factor of $1$ at each step (except the last):
\[
  \begin{array}{rcl}
    F(n + 1) & = & 1 · F(n) + F(n - 1),     \\
    F(n)     & = & 1 · F(n - 1) + F(n - 2), \\
             & ⋮ &                          \\
    F(2)     & = & 2 · F(1).
  \end{array}
\]
Other numbers may have higher factors -- for example, the silver ratio has a
factor of $2$ at each step -- but the Fibonacci numbers have the smallest factor.

This property is also reflected in the fractions $a/b$ for a given input $(a, b)$.
At each step (except the last) the fraction always has an integer part of $1$:
\[
  \begin{array}{rcl}
    \frac{F(n + 1)}{F(n)} & = & 1 + \frac{F(n - 1)}{F(n)},     \\
    \frac{F(n)}{F(n - 1)} & = & 1 + \frac{F(n - 2)}{F(n - 1)}, \\
                          & ⋮ &                                \\
    \frac{F(2)}{F(1)}     & = & 2.                              \\
  \end{array}
\]
So, a higher-dimensional Fibonacci number would also need to have a fraction
with integer part $1$ at each step.

% TODO: We should also discuss the thing about idempotence under the choice of the pivot.
% TODO: Should we have a discussion about why we choose this particular linear system.

First, we are indeed talking about just numbers and not vectors.
The Euclidean algorithm takes one-dimensional integers as input, so one would
expect that the generalized algorithm would take Fibonacci vectors as input.
But the numbers of steps that the algorithm requires only depends on the
initial solution vector x.
In fact, the entire execution of the algorithm only depends on this vector.
There are infinitely many possible linear systems which have the same solution
vector x, and hence there would be infinitely many "Fibonacci vectors".
Therefore, we look instead only for the numbers inside the rational vector.

There are many different possibilities for choosing a linear system which has the solution $x$.
We use a simple construction, which
given a rational vector $x = \left(\frac{p₁}{q}, \frac{p₂}{q}, …, \frac{p_d}{q}\right)$,
constructs the linear system $Bx = c$ with
\begin{equation}
  \label{eq:linear-system-construction}
  B =
  \begin{pmatrix}
    q & 0 & ⋯ & 0 \\
    0 & q & ⋯ & 0 \\
    ⋮ & ⋮ & ⋱ & ⋮ \\
    0 & 0 & ⋯ & q \\
  \end{pmatrix},
  \quad
  c =
  \begin{pmatrix}
    p₁ \\
    p₂ \\
    ⋮ \\
    p_d
  \end{pmatrix}.
\end{equation}

\begin{table}[tbp]
  \caption{The first 10 Fibonacci numbers for $d = 1, …, 5$ and their respective golden ratio.}
  \label{tbl:min-fibonacci}
  \centering
  \input{tables/min-fibonacci}
\end{table}

% I would like a better explanation of what the goal of this section is

Although we have reduced the problem just to the vector $x$, there is still
some ambiguity left, when extending Fibonacci numbers to higher dimensions.
The generalized Euclidean algorithm allows for an additional degree of freedom
by choosing which column vector $B_ℓ$ we swap our vector $c$ with.
Therefore, there cannot be just one definitive Fibonacci number, there are
different ones depending on the choice of our index $ℓ$.

The first strategy one may think of is to choose the index $ℓ$ with the
smallest fractional value.
After all, this gives us the largest decrease in the determinant over one
iteration.
Locally, this would be the highest decrease we can achieve.
So we begin with this strategy and try to find a generalization of Fibonacci
numbers under this strategy.

% TODO: Actually explain how we derive the Fibonacci numbers for this strategy.
% TODO: Should we have a table for the Fibonacci numbers?
% TODO: What about adding references to their OEIS number?

The Fibonacci numbers for this strategy are defined as
\begin{align*}
  F(0) = F(1) = ⋯ = F(d) = 1, \qquad F(n + 1) = F(n) + F(n - 1) + ⋯ + F(n - d).
\end{align*}
For $d = 2, 3, 4$, they are also called the Tribonacci, Tetranacci and Pentanacci numbers.
In general, they are known as the $d$-bonacci numbers.
For these numbers,
We use the solution vector
\[
  x =
  \left(
    \frac{F(n)}{F(n + 1)},\;
    \frac{F(n) + F(n - 1)}{F(n + 1)},\;
    ⋯,\;
    \frac{F(n) + F(n - 1) + ⋯ + F(n + 1 - d)}{F(n + 1)}\;
  \right)
\]
and we can use the construction from Equation~\ref{eq:linear-system-construction}
for the actual input to the generalized Euclidean algorithm.

The smallest fractional value in this vector is $x₁$, so we pivot with $ℓ = 1$ first.
\begin{align*}
  \{x₁'\}
  & = \left\{\frac{1}{\{x₁\}}\right\}
  = \left\{\frac{F(n + 1)}{F(n)}\right\}
  = \frac{F(n - 1) + ⋯ + F(n - d)}{F(n)}.
\end{align*}
This gives us the value for $x_d$, if we would have started with $F(n)$ instead of $F(n+1)$.
The other values in our input vector $x$ are calculated as follows:
\begin{align*}
  \{xᵢ'\}
  & = \left\{\frac{\{xᵢ\}}{\{x₁\}}\right\} \\
  & = \left\{\frac{F(n) + F(n - 1) + ⋯ + F(n + 1 - i)}{F(n + 1)} · \frac{F(n + 1)}{F(n)}\right\} \\
  & = \frac{F(n - 1) + ⋯ + F(n - d)}{F(n)}.
\end{align*}
So in the next iteration the value $x_i$ has the same value as $x_{i+1}$ if we
would have started with $F(n)$.
Therefore, in the next iteration the smallest fractional value must be $x_2$.
We continue this until we have cycled through all variables.

Another example would be the opposite strategy: Choosing the maximum in each iteration.
In this case, the Fibonacci numbers are defined as
\begin{align*}
  F(0) = \dots = F(d) = 1, \qquad F(n + 1) = F(n) + F(n - d).
\end{align*}

The golden ratios for each definition is the limit of consecutive Fibonacci numbers
\[
  φ = \lim_{n → ∞} \frac{F(n + 1)}{F(n)},
\]
where $F(n)$ can be any of the previously defined Fibonacci numbers.

% ==============================================================================
\iffalse
\section{Higher-Order Fibonacci Sequence for the Maximum Strategy}
\fi
% ==============================================================================

\begin{table}[tbp]
  \caption{The first 10 Fibonacci numbers for $d = 1, …, 5$ and their respective golden ratio.}
  \label{tbl:max-fibonacci}
  \centering
  \input{tables/max-fibonacci}
\end{table}

Next, we look at the opposite strategy:
Always taking the index with the maximum fractional value at each step.

The Fibonacci numbers are defined by the following recurrence relation:
\begin{align*}
  F(0) = F(1) = \dots = F(d) = 1, \qquad F(n + d + 1) = F(n + d) + F(n).
\end{align*}
The respective golden ratio $φ$ for this sequence is the positive root of the polynomial
\begin{align*}
  p(x) = x^{d+1} - x^d - 1.
\end{align*}

The golden ratio for this sequence also shows up for a different strategy.
When we choose two indices $ℓ$ and $ℓ'$ such that they minimize the ratio
$\{x_{ℓ'}\}/\{x_{ℓ}\}$, then we can bound the decrease of the determinant over
two iterations using the golden ratio $φ$.

% TODO: I feel like this was using $ψ = φ^{-1}$ instead of $φ$.
\begin{theorem}
  The determinant decreases by at least $φ^{d+1}$ over two iterations with $d ≥ 2$.
\end{theorem}

\begin{proof}
  We assume WLOG that the vector is sorted in increasing order and
  $0 ≤ x_i ≤ 1$ for every $i ≤ d$.
  For a contradiction, assume the algorithm yields a smaller decrease than $φ^{d+1}$ on input $x$.
  We must have $x_i > φ^{d+1-i}$ for every $i = 1, \dots, d$.
  For the first value, we have $x₁ > φ^d$, because otherwise we have a total decrease of
  \[
    x₁ \left\{ \frac{x₂}{x₁} \right\} ≤ φ^{d-1} - φ^d = φ^{2d+1} < φ^d.
  \]
  Suppose $x_i > φ^{d+1-i}$ and $x_{i+1} ≤ φ^{d-i}$.
  Then we can achieve a total decrease of
  \[
    x_i · \left\{ \frac{x_{i+1}}{x_i} \right\} ≤ x_{i+1} - x_i < φ^{d-i} - φ^{d+1-i} = φ^{2d+1-i} < φ^d.
  \]
  It follows that $x_i > φ^{d+1-i}$ for every $i ≤ d$.
  But then
  \[
    x_d \left\{ \frac{1}{x_d} \right\} ≤ 1 - x_d < 1 - φ = φ^{d+1}.
  \]
  Hence, we achieve a decrease of at least $φ^{d+1}$ over two iterations.
\end{proof}

The bound for this strategy is tight.
We can construct an input that moves arbitrarily close to this bound.
Formally, for every sufficiently small $ε > 0$, we can find an input $x$ which
achieves a decrease of at most $φ^{d+1} - ε$.
The idea is to choose $x_d$ to be just over $φ$ and all other variables as a multiple of $x_d$
such that we have to choose $x_d$.
But choosing $x_d$ only gives us a decrease of $φ^{d+1} - ε$ in total.

% TODO: Explain what sufficiently small means
\begin{theorem}
  For every (sufficiently small) $ε > 0$,
  there exists an input $x ∈ ℝ^d$ with $d ≥ 2$,
  which achieves a decrease in the determinant of exactly $φ^{d+1} - ε$ over two
  iterations.
\end{theorem}

\begin{proof}
  We choose $x_i = φ^{d+1-i} (1 + ε)$ for $i = 1, \dots, d$.
  The strategy chooses between
  \[
    \left\{ \frac{x_{i+1}}{x_i} \right\}
    = \frac{φ^{d-i} (1 + ε)}{φ^{d+1-i} (1 + ε)} - 1 = \frac{1}{φ} - 1,
    \text{ and }
    \left\{ \frac{1}{x_d} \right\}
    = \frac{1}{φ + ε} - 1
  \]
  Clearly, the strategy chooses $x_d$ since its ratio $\{1/x_d\}$ is the smallest.
  Therefore, the total decrease over two iterations is
  \[
    (φ + ε) \left\{ \frac{1}{φ + ε} \right\} = 1 - φ - ε = φ^{d+1} - ε.
    \qedhere
  \]
\end{proof}

% begin periodicity
Given, the linear recurrence $F(n + d + 1) = F(n) + a_1 F(n - 1) + \dots + a_d F(n + d)$,
the value for $x_i^{(n)}$ in the solution $x^{(n)}$ is
\begin{equation}
  \label{eq:general-solution}
  x_i^{(n)} = \frac{F(n - i) + \sum_{k=1}^{i-1} a_{d-k} F(n - k)}{F(n)}.
\end{equation}

\begin{lemma}
  Pivoting with the first element of $x^{(n)}$ using the modified update rule yields the vector
  \[
    x' = (x^{(n-1)}_2, x^{(n-1)}_3, \dots, x^{(n-1)}_d, x^{(n-1)}_1).
  \]
\end{lemma}

\begin{proof}
  For $x_1$, we get
  \[
    \begin{aligned}
      \frac{1}{x_1}
      & = \frac{F(n)}{F(n - 1)} \\
      & = a_d + \frac{F(n - d - 1) + a_1 F(n - d - 2) + \dots + a_{d-1} F(n - 2)}{F(n - 1)} \\
      & = a_d + x^{(n-1)}_d.
    \end{aligned}
  \]
  For the other values, we get
  \begin{align*}
    \frac{x_i}{x_1}
    & = \frac{F(n - i) + \sum_{k=1}^{i-1} a_{d-k} F(n - k)}{F(n)} \frac{F(n)}{F(n - 1)} \\
    & = a_{d-1} + \frac{F(n - i) + \sum_{k=2}^{i-1} a_{d-k} F(n - k)}{F(n-1)} \\
    & = x^{(n-1)}_{i+1} \qedhere
  \end{align*}
\end{proof}

\begin{corollary}
  Running the generalized Euclidean algorithm with $x^{(n)}$ as the solution to
  the linear system $B x = c$ requires $n$ steps, if $x^{(n)}_1$ is the
  smallest element in $x^{(n)}$.
\end{corollary}

Of course, the Euclidean algorithm receives a matrix $B \in \Z^{d \times d}$
and vector $c \in \Z^d$ as its input and not the solution $x$ itself.
However, we can construct a very simple linear system $B^{(n)} x = c^{(n)}$,
where $x^{(n)}$ is the solution, in the following way:
\[
  B^{(n)} = F(n) I_d, \qquad c^{(n)}_k = \sum_{i=0}^k a_{d-i} F(n - i) \text{ for } k ≤ d.
\]

\begin{lemma}
  The vector $x^{(n)}$ from Equation~\ref{eq:general-solution} is a solution to the
  linear system $B^{(n)} x = c^{(n)}$.
\end{lemma}

\begin{proof}
  Each row in the linear system is of the form
  \[
    F(n) x_k = \sum_{i = 0}^k a_{d-i} F(n - i).
  \]
  Rearranging the equation gives the definition of $x^{(n)}_k$.
\end{proof}
% end periodicity

% ==============================================================================
\section{Multi-Dimensional Golden and Metallic Ratios}
% ==============================================================================

% begin convergence
Dividing two consecutive Fibonacci numbers approaches the golden ratio as $n$ increases.
The golden ratio is a solution to the equation $x^2 - x - 1 = 0$.
For higher dimensions, we will encounter similar polynomial equations with higher degree.
The goal of this section is to generalize the relationship between linear
recurrences like the Fibonacci sequence with their respective golden ratio.

\begin{definition}
  A linear recurrence with coefficients~$a_0, \dots, a_d$ is an equation of the
  following form:
  \[
    F(n + 1) = a_0 F(n - d) + a_1 F(n - d + 1) + \dots + a_{d-1} F(n - 1) + a_d F(n).
  \]
\end{definition}

\begin{definition}
  Given a linear recurrence~$F$ with coefficients~$a_0, \dots, a_d$, its
  \emph{characteristic polynomial} $p_F$ is defined as
  \[
    p_F(x) = x^{d+1} - a_d x^d - a_{d-1} x^{d-1} - \dots - a_1 x - a_0.
  \]
\end{definition}

\begin{example}
  The linear recurrence which has $p_d^*(x)$ as its characteristic polynomial is:
  \begin{align*}
    F(n) = F(n - 1) + (-1)^{d+1} F(n - d) - \sum_{i=2}^{d - 1} (-1)^{i+1} 2 F(n - i).
  \end{align*}
\end{example}

Next, we will show that the ratios of consecutive terms
\[
  r_n = \frac{F(n+1)}{F(n)} = \frac{a_0 F(n - d)}{F(n)} + \frac{a₁ F(n - d + 1)}{F(n)} + ⋯ + \frac{a_{d-1} F(n-1)}{F(n)} + 1.
\]
is converging to the only positive real root of the characteristic polynomial $p_F$.
Each ratio in the sum can be rewritten as a product of consecutive ratios $r_{n-i}$:
\begin{align*}
  \frac{F(n - d + i)}{F(n)}
  & = \frac{F(n - d + i + 1)}{F(n - d + i)} \frac{F(n - d + i + 2)}{F(n - d + i + 1)} \dots \frac{F(n-1)}{F(n)} \\
  & = \frac{1}{r_{n - d + i}} · \frac{1}{r_{n - d + i + 1}} · \dots · \frac{1}{r_{n-1}}.
\end{align*}
So we can calculate the ratio $r_n$ using the previous ratios $r_{n-1}, r_{n-2}, …, r_{n-d}$ as follows:
\[
  r_n = 1 + \frac{a_d}{r_{n-1}} + \frac{a_{d-1}}{r_{n-1} r_{n-2}} + ⋯ + \frac{a₀}{r_{n-1} r_{n-2} \dots r_{n-d}}.
\]


\begin{lemma}
  The ratios $r_n$ are bounded between $1$ and $d+1$.
\end{lemma}

\begin{proof}
  The first $d - 1$ ratios are all $1$.
  The ratio $r_d$ is equal to
  \[
    \frac{F(d+1)}{F(d)} = \frac{F(0) + F(1) + ⋯ + F(d)}{F(d)} = \frac{1 + 1 + ⋯ + 1}{1} = d + 1,
  \]
  which also satisfies the bounds.
  By induction, suppose that the ratios $r_{n-1}, r_{n-2}, …, r_{n-d}$ all satisfy the bound between $1$ and $d+1$.
  The ratio $r_n$ can be reformulated as follows:
  \[
    r_n = 1 + \frac{a₁}{r_{n-1}} + \frac{a₂}{r_{n-1} r_{n-2}} + \dots + \frac{a_d}{r_{n-1} r_{n-2} \dots r_{n-d}}.
  \]
  Since $r_{n-i} ≤ d+1$, we can bound $r_n$ from below by
  \[
    r_n ≥ 1 + \frac{a₁}{d+1} + \frac{a₂}{(d+1)^2} + \dots + \frac{a_d}{(d+1)^d} ≥ 1
  \]
  and since $r_{n-i} ≥ 1$, we can bound $r_n$ from above by
  \[
    r_n ≤ 1 + \frac{a₁}{1} + \frac{a₂}{1} + \dots + \frac{a_d}{1} ≤ d+1.
  \]
  Hence, $1 ≤ r_n ≤ d+1$ for every $n ≥ 0$.
\end{proof}

The next step to prove the convergence of the ratios would be to show that the
ratios are monotonically increasing or decreasing.
That is clearly not the case, however.
For $d=1$, the only sequence is the Fibonacci sequence and its ratios are
always alternating between increasing and decreasing.
But they do approach the golden ratio.
For the convergence of the generalized sequence $r_n$,
we will instead apply the squeeze theorem.
The main idea is illustrated in Figure~\ref{fig:fibonacci-convergence}.
We show that the sequences
\[
  s_n = \min\{r_n, r_{n-1}, …, r_{n-d} \}, \qquad t_n = \max\{r_n, r_{n-1}, …, r_{n-d}\}
\]
converge to the same limit.
Because $r_n$ lies between the two sequences, it must also converge to the same limit as $s_n$ and $t_n$.
For these sequences, we already know from the previous lemma that they are bounded,
so it only remains to be shown that they are increasing and decreasing,
respectively.

% TODO: Include a plot of the Fibonacci numbers

\begin{lemma}
  The sequences $s_n$ and $t_n$ are monotone.
\end{lemma}

\begin{proof}
  Each ratio can be represented as a convex combination of the previous ratios, i.e.
  \[
    r_{n+1} = λ₀ r_n + λ₁ r_{n-1} + \dots + λ_d r_{n-d}
  \]
  using $λ_i = F(n - i) / F(n + 1)$.
  To show that this is indeed a convex combination, all coefficients $λ_i$
  need to be non-negative and $λ₀ + λ₁ + \dots + \lambda_d = 1$.
  The former follows from the fact that Fibonacci numbers are always increasing,
  while the latter follows simply from the definition of the Fibonacci numbers.
  Because $s_n$ is the minimum and $t_n$ the maximum of $r_n, r_{n-1}, …, r_{n-d}$,
  we can bound the next maximum by
  \[
    t_{n+1} ≤ r_{n+1} = λ₀ r_n + λ₁ r_{n-1} + \dots + λ_d r_{n-d} ≤ λ₀ t_n + λ₁ t_n + ⋯ + λ_d t_n = t_n.
  \]
  and the next minimum by
  \[
    s_{n+1} ≥ r_{n+1} = λ₀ r_n + λ₁ r_{n-1} + \dots + λ_d r_{n-d} ≥ λ₀ s_n + λ₁ s_n + ⋯ + λ_d s_n = s_n.
  \]
  Therefore, $s_n ≤ s_{n+1} ≤ t_{n+1} ≤ t_n$.
\end{proof}

\begin{lemma}
  The sequences $s_n$ and $t_n$ are converging to the same limit.
\end{lemma}

% TODO: Finish this proof
\begin{proof}
  For a contradiction, suppose there exists some $δ > 0$ such that $t_n - s_n > δ$ for every $n ≥ 0$.
  Out of the previous ratios, there must be one ratio $r_k$ exactly equals to $t_n$.
  Therefore,
  \begin{align*}
    s_{n+1} ≥ r_n & = λ₀ r_{n-d-1} + λ₁ r_{n-d} + ⋯ + λ_d r_{n-1} \\
                  & ≥ λ_k t_n + \sum_{i ≠ k} λ_i s_n \\
                  & = λ_k t_n + (1 - λ_k) s_n = s_n + λ_k (t_n - s_n) ≥ s_n + λ_k δ.
  \end{align*}
  We have
  \[
    λ_k = \frac{F(n+k)}{F(n+d+1)} = \frac{1}{r_{n+k} r_{n+k+1} \dots r_{n+d+1}} ≥ \frac{1}{(d+1)^{d+1-k}}.
  \]
  Hence, $λ_k$ is always greater than some constant $c > 0$ for every $k ≥ 0$.
  But then
  \[
    s_{n+i} ≥ s_{n+i-1} + c δ ≥ s_{n+i-2} + 2c δ ≥ \dots ≥ s_n + i c δ
  \]
  and $s_{n+i}$ would increase indefinitely as $i$ approaches infinity,
  which contradicts the previous lemma.
  Therefore, $δ = 0$ and it follows that $s_n$ and $t_n$ are approaching the same limit.
\end{proof}

% TODO: Show that this converges! We're still missing the convergence criteria
\begin{theorem}
  Let $F$ be a linear recurrence with coefficients $a_0, \dots, a_d ∈ \{0,1\}$
  and let $φ$ be the real positive root of its characteristic polynomial.
  Then,
  \[
    \lim_{n \to \infty} \frac{F(n + 1)}{F(n)} = φ.
  \]
\end{theorem}

\begin{proof}
  By the previous lemma, the ratios $r_n$ approach some limit $r$. It follows:
  \[
    r
    = \lim_{n → ∞} r_n
    = \lim_{n → ∞} 1 + \frac{a_d}{r_{n-1}} + \frac{a_{d-1}}{r_{n-1} r_{n-2}} + ⋯ + \frac{a₀}{r_{n-1} r_{n-2} \dots r_{n-d}}.
  \]
  Hence, each denominator in the sum approaches $r^i$,
  which results in the following polynomial:
  \[
    r = a₀ + \sum_{i = 1}^d \frac{a_{d - i}}{r^i}
    \iff
    r^{d+1} = a₀ + a₁ r + \dots + a_d r^d,
  \]
  which directly corresponds to a root of its characteristic polynomial.
  Furthermore, the ratios are always positive, so $r = φ$.
\end{proof}

\begin{corollary}
  For the same recurrence $F$, $\lim_{n → ∞} F(n + i) / F(n) = φ^i$ for $i > 1$.
\end{corollary}

% end convergence

\begin{definition}
  Given a polynomial $p = \sum_{i=0}^n a_i x^n$, its \emph{reciprocal polynomial},
  denoted as $p^*$, is defined as
  \[
    p^*(x) = x^n p(x^{-1}) = a_n + a_{n-1} x + \dots + a_0 x^n.
  \]
\end{definition}

\begin{example}
  The reciprocal of the golden ratio polynomial $x^2 - x - 1$ is
  \begin{align*}
    x^2 + x - 1,
  \end{align*}
  which has a root at $ψ = 1/φ$.
\end{example}

\begin{lemma}
  Let $p$ be a polynomial. Then, $a$ is a root of $p$ if and only if $a^{-1}$ is a root of $p^*$.
\end{lemma}

\begin{proof}
  Let $a$ be a root of $p$. It follows $p^*(a^{-1}) = a^n p(a) = a^n \cdot 0 = 0$.
  Now let $a^{-1}$ be a root of $p^*$. Then $p^*(a^{-1}) = a^n p(a) = 0$.
  By construction, $a$ cannot be zero and therefore we must have $p(a) = 0$.
\end{proof}

% ==============================================================================
\iffalse
\section{Metallic Ratios}
\fi
% ==============================================================================

The result from the previous section can be generalized to an even larger class
of polynomials by allowing $a_1, a_2, …, a_d ∈ \Z_{≥ 0}$ instead of only binary
values.
The roots of these polynomials can be seen as a generalization of the metallic ratios.
Indeed, for the one-dimensional case the polynomials are
\[
  x^2 - a₁ x - 1,
\]
which characterizes all possible metallic ratios, like the silver ratio ($a₁ = 2$) or
the bronze ratio ($a₁ = 3$).
The generalization considers higher-degree polynomials of the form
\[
  x^{d+1} - a_d x^d - \dots - a_1 x - 1
\]
with non-negative integer coefficients $a_1, a_2, …, a_d$.
Each of these polynomials has a corresponding linear recurrence,
which can be reconstructed by the characteristic polynomial.
For the linear recurrence,
the ratios of consecutive terms approaches the real positive root of the
polynomial again.
